A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-09-09 jacobian_calc_CMAME.tex
Results and conslusion sections

\subsection{The Comparative Study}

The methods were compared by running two sets of test problems where each method would solve a
problem concurrently. Each test problem simulated a $4 by .25 by .25$ meter block of metal
undergoing tension along the axis parallel to the long dimension of the block. The test problems
were set up in \emph{Perdigm} as qausistatic equillibrium problems, where the displacement boundary
conditions used to apply tension were applied gradually, in 'load steps', and an equilibrium
solution for each load step was achieved before applying the next load step. The purpose of applying
the load in gradual steps is a widely used technique to ensure the Newton-Raphson method is always
guessing near the actual solution. The material properties of the model were set at $8.0E-9 metric
tonne/ mm^3$ density, $1.515E4 MPa$ bulk modulus, and $7.813E4 Mpa$ shear modulus. The peridynamic
horizon used in a test problem was three times the node spacing for the corrisponding model mesh.
This value for horizon was chosen out of experience to insure that every node could have at least
one nonlocal connection along the diagonal of the node grid.

@@

\begin{center}
    \begin{tabular}{c c c c }
    \hline
    \multicolumn{4}{c}{Averaged Results, Entire Test Matrix} \\ \hline
    Cores & Jacobian Els. & Calc. Time (s) & Accuracy Diff. (MPa)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
	1 & 1.99E6  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{center}

\subsection{Speed Data}

@@

@@

\subsection{Accuracy Data}

\begin{figure}[h]
  \centering
  \scalebox{.7}{\input{./figs/serial_accuracy.pgf}}
  \caption{Serial test series speed measurements.}
  \label{fig:serial_accuracy}
\end{figure}

@@

\subsection{Efficiency Data}

@@

@@

%%\includegraphics[width=\textwidth]{serial_speed_rel.png}
%%\includegraphics[width=\textwidth]{multi_speed_rel.png}

@@

\subsection{Thoughts on Results}

@@

After the bulk of the study had been conducted, it was determined that the implementations of
forward-difference, central-difference and by virtue of reusing code, complex-step, in \emph{Peridigm} were
not optimal. Each of these implementations included extensive looping over noncontiguous addresses
in memory and unnecessary re-calculation of constant values. The approach used does minimize
temporary memory use, but is very wasteful of cpu and memory bandwidth. Unpublished preliminary results show
that when calculations for the quantity 'zeta' are re-used and only perturbed and previously
pertubed functions of independent variables are recomputed throughout the course of a Jacobian
evaluation, complex-step performs at the speed of unmodified forward-difference. Because of this it is
speculated that forward-difference and central-difference could be modified in nearly the same way to
perform more quickly than AD. An enhanced performance implementation of forward-difference will
eventually be made available on the corresponding author's website.

\subsection{Final Thoughts on Complex-Step} Complex-step was shown to be highly accurate by the
standards of this study, but as it is implemented in \emph{Peridigm} may not be any better than AD,
and it is almost certainly slower for a naive implementation. When looking to the future of
\emph{Peridigm}, Complex-step holds the advantage of relying on byte-copiable datatypes, while AD
requires a complicated serialization and deserialization in order to function as byte-copiable. This
is fine for copying in between MPI processes, but deserialization on an accelerator or GPU is very
difficult, as this would require compiling a version of the AD library being used that were
compatible with the accelerator or GPU being used. What this means is that if \emph{Peridigm} were
to be reformulated to take advantage of accelerators or GPUs, attractive given their performance and
power efficiency advantages over traditional CPUs, the flexibilty of complex-step owed to its simple
implementation may make it more viable fo rthat application.

>>>>

\subsection{The Comparative Study} 
\label{tcs}

The methods were compared by running two sets of test problems where each method would solve a
problem concurrently. Each test problem simulated a $4 by .25 by .25$ meter block of metal
undergoing tension along the axis parallel to the long dimension of the block. The test problems
were set up in \emph{Perdigm} as qausistatic equillibrium problems, where the displacement boundary
conditions used to apply tension were applied gradually, in 'load steps', and an equilibrium
solution for each load step was achieved before applying the next load step. The purpose of applying
the load in gradual steps is a widely used technique to ensure the Newton-Raphson method is always
guessing near the actual solution. The material properties of the model were set at $8.0E-9 metric
tonne/ mm^3$ density, $1.515E4 MPa$ bulk modulus, and $7.813E4 MPa$ shear modulus. The peridynamic
horizon used in a test problem was three times the node spacing for the corrisponding model mesh.
This value for horizon was chosen out of experience to insure that every node could have at least
one nonlocal connection along the diagonal of the node grid.

@@

This section contains the averaged results from the study presented in tabular and graphic format.
Table \ref{tab:results} shows the core count, number of nonzero Jacobian matrix elements, load step
average calculation time and load step average accuracy. The units for accuracy are derived from the
expression used to calculate the Jacobian matrix elements and the units used to define bulk material
properties, mentioned in \ref{tcs}.

\begin{table}
    \begin{tabular}{c c c c }
        \caption{Averaged Results, Each Test} \label{tab:results}
    \hline
    Cores & Jacobian Els. & Calc. Time (s) & Accuracy Diff. (MPa)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
	1 & 1.99E6  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{table}

The data presented in \ref{tab:results} will now be shown in graphic form alongside some discussion
of results and trends.

\subsection{Speed Data}
Speed measurements were taken according to the methods mentioned in \ref{tcs}. The averaged results
of these speed measurments for the single core tests can be seen in \ref{fig:serial_speed}.
Average time in seconds to compute a Jacobian matrix element is plotted against number of nonzero
Jacobian matrix ekements.
Complex-step turned out to be the slowest of the four methods, taking the longest average time per
iteration. The next slowest method was central-difference, followed by forward-difference and
automatic-differentiation. Because the x axis of \ref{fig:serial_speed} is on a log scale, it
bears mentioning that while visually it may appear that the trend indicated by the progression of
datapoints on the plot is non-linear, the relationship would be closer to linear. This bodes well
for the scalability of each of the methods by themselves in \emph{Peridigm}.

@@

For the multicore tests, averaged speed results appear in \ref{fig:multi_speed}. For this series,
rather than varying the number of nonzero Jacobian matrix elements the number of cores used to
solve the problem is varied, while the number of nonzero Jacobian matrix elements remained constant.
Complex-step again was the slowest method, followed by central-difference, automatic-differentiation
and then forward-difference, taking the title of fastests method for this group of tests.
\ref{fig:multi_speed} indicates that when looking at average in-iteration speed, each of the
methods scale linearly with the number of cores used for the given conditions of this group of
tests. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale
well when additional processors are used. An explanation as to why automatic-differentiation and
forward-differences swapped places in speed ranking relative to the single core results was not
determined.

@@

\subsection{Accuracy Data}
Accuracy measurements where taken according to the methods mentioned in \red{tcs}. The averaged
results of these accuracy measurements for the single core tests can be seen in
\ref{fig:serial_accuracy}. Averaged Frobenius norm of the element-wise difference between a given
Jacobian and the Jacobian produced by the automatic-differentiation method is plotted against number
of nonzero Jacobian matrix elements. Complex-step was revealed as the most accurate of the methods
besides automatic-differentiation, complex-step resulting in several orders of magnitude less
difference that resulted from using the central-difference method and the most inaccurate
forward-difference method. This accuracy ranking order was mantained for each test in the series. It
should be pointed out compared to the L2 norms of any of the Jacobian matrices by themselves, the
magnitude of these differences for any of the methods is relatively small. These values appear in
the experimental data which is available on the corresponding author's website. The effect of these
accuracy differences on convergence rate directly was not determined.

\begin{figure}[h]
  \centering
  \scalebox{.7}{\input{./figs/serial_accuracy.pgf}}
  \caption{Serial test series accuracy measurements.}
  \label{fig:serial_accuracy}
\end{figure}

The averaged resuts of the accuracy measurements taken for the multicore test series can be seen in
\ref{fig:multi_accuracy}. Y axis units remain unchanged from \ref{fig:serial_accuracy}, but the x
axis now indicated number of cores used to solve the test problem. Accuracy trends matched those for
the single core tests.

@@

\subsection{Efficiency Data}
A derived measurement was also used to compare the methods. The single core results are shown in
\ref{fig:serial_efficiency}. Efficiency, calculated as the average number of seconds taken per nonzero Jacobian matrix element 
was ploted against the number nonzero Jacobian matrix elements. The ranking order of the methods
matched their speed ranking in \ref{fig:serial_speed}. This plot indicates that there is a small,
yet present decrease in efficiency w.r.t number of nonzero Jacobian elements. But judging by the
speed results, this trend in efficiency doesn't appear to have great effect in human terms for the
mesh densities used in this test series.

@@

The multicore efficiency results are shown in \ref{fig:multi_efficiency}. Y axis is average number
of seconds taken per nonzero Jacobian matrix element times number of cores used to solve the
problem. The x axis in number of cores used to solve the problem. The multicore efficiency plot
shows that automatic-differentiation was slower than forward-difference because for the given test
problem it had a lower efficiency. Strangely automatic-differentiation appeared to gain in
efficiency over the last few datapoints in the plot, looking at the raw data confirmed that this was
what was measured rather than this being an error during data reduction. The reason behind this
trend was not determined.

@@

@@

\subsection{Thoughts on Results} Based on the experimental results, it appears that for users of
\emph{Peridigm} there is a case to be made for using automatic-differentiation for reasons of
performance and accuracy. Although, the greater speed of finite-difference for certain multicore
tests done in this study may give some users pause. While the results showed that complex-step was
clearly more accurate than central-difference and forward-difference for the parameters of the
tests, it was not determined whether or not this is a clear advantage of complex-step over
finite-difference in terms of accuracy of final results and speed of convergence.   For other
simulation software it is obviously necessary to collect the same types measurements as decribed in
this paper in order to make a design decision. For new developments it is recommended that combined
accuracy and speed of iteration mesurments be conduted along with entirely separate convergence
studies.  For reasons explained in the next subsection, a convergence study in \emph{Peridigm} performed using
the implementations of the methods used for this paper would be moot.

@@

After the bulk of the study had been conducted, gained familiarity with \emph{Peridigm} revealed
that the implementations of forward-difference, central-difference and by virtue of reusing code,
complex-step, in \emph{Peridigm} were not optimal due to inefficiencies in the implementations of
the Linear Peridynamic Solid (LPS) model. Each of these implementations included extensive looping
over noncontiguous addresses in memory and unnecessary re-calculation of constant values.  The
approach used does minimize temporary memory use, and preserve modularity of the code for non-LPS
models, but is very wasteful of cpu and memory bandwidth for use with LPS. Unpublished preliminary
results show that when calculations for the quantity 'zeta' are re-used and only perturbed and
previously pertubed functions of independent variables are recomputed throughout the course of a
Jacobian evaluation, complex-step performs at the speed of unmodified forward-difference. Because of
this it is speculated that forward-difference and central-difference could be modified in nearly the
same way to perform more quickly than AD. In fact, because forward difference involves double the
number of model evaluations than that of complex-step, forward-difference stands to gain
proportionally more efficiency. Although no formal academic paper will be likely written based on
this, a brief report and an enhanced performance implementation of forward-difference for peridynamic LPS
in \emph{Peridigm} will eventually be made available on the corresponding author's website and is
intended to be submitted for official inclusion in future releases of \emph{Peridigm}.

\subsection{Final Thoughts on Complex-Step} Complex-step was shown to be highly accurate by the
standards of this study, but as it is implemented in \emph{Peridigm} may not be any better than AD,
and it is almost certainly slower for a naive implementation. When looking to the future of
\emph{Peridigm}, Complex-step holds the advantage of relying on byte-copiable datatypes, while AD
requires a complicated serialization and deserialization in order to function as byte-copiable. This
is fine for copying in between MPI processes, but deserialization on an accelerator or GPU is very
difficult, as this would require compiling a version of the AD library being used that were
compatible with the accelerator or GPU being used. What this means is that if \emph{Peridigm} or any
simulation software were to be reformulated to take advantage of accelerators or GPUs, attractive
given their performance and power efficiency advantages over traditional CPUs, the flexibilty of
complex-step owed to its simple implementation may make it more viable fo rthat application.
